\begin{abstract}

Fluid simulation is an extremely common and important computational task. These simulations are often heavily expensive and require a large amount of CPU time, hence difficult to apply in real time applications. Fortunately, modern GPUs, armed with massively parallel general purpose computing architectures, provide a solution to this problem. This project explores methods to perform fluid simulations on GPUs, and to render the results of the simulations in a visually plausible manner, all in real time. 


This project focuses on three of the most widely used fluid simulation algorithm: FLIP (Fluid-Implicit-Particle), PBF (Position Based Fluids), and PCISPH (Predicative-Corrective Incompressible Smoothed Particle Hydrodynamics). The project looks into how each algorithm can be adapted to perform well on modern GPUs, and creates efficient GPU implementations of these algorithms using NVIDIA's CUDA programming language. The project also extends the FLIP algorithm to support multiple fluid rendering, thereby capturing the diffusion phenomenon between different phases of fluids. Alongside the simulation, the project also implemented a real time liquid rendering scheme, which efficiently captures the refection and refraction effects, as well as the varying concentrations of colored fluid phases inside the liquid. The final result takes the form of a fully functional software, which performs and renders liquid simulations in real time, that can be run on machines with recent NVIDIA GPUs.

\end{abstract}
